A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

To represent infinite values such as $\const{S}(\const{S}(\ldots))$, a natural
choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. The verified C compiler CompCert \cite{leroy-2009},
the verified Java compiler JinjaThreads \cite{lochbihler-2010-jinja}, and the
formalized Java memory model \cite{lochbihler-2014-jmm} all rely on
codatatypes to capture infinite processes.

Codatatypes, like datatypes, are freely generated by their constructors. But
whereas datatype values are finite, for codatatypes infinite constructor terms
also denote values. Intuitively, a codatatype consists of all finite and
infinite ground constructor terms, and only those. For example, the
specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat})\]
%
(expressed in a notation inspired by typed functional programming languages)
introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{ES}(\const{ES}(\const{ES}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat}) \,\mid\, \const{EInf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{ES}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{ES}(\const{ES}(\const{ES}(\ldots)))$.

>>>>

To represent infinite values such as $\const{S}(\const{S}(\ldots))$, a natural
choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. The verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler JinjaThreads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely on codatatypes to capture infinite
processes. Many proof assistants, including Agda, Coq, Isabelle, Matita, and
PVS, explicitly support them, as do the CoALP dialect of Prolog and the Dafny
specification language.

Codatatypes, like datatypes, are freely generated by their constructors. But
whereas datatype values are finite, for codatatypes infinite constructor terms
also denote values. Intuitively, a codatatype consists of all finite and
infinite ground constructor terms, and only those. For example, the
specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{ES}(\const{ES}(\const{ES}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat}) \,\mid\, \const{EInf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{ES}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{ES}(\const{ES}(\const{ES}(\ldots)))$.
